Meta-analytic approaches and effect sizes to account for 'nuisance heterogenity' in comparative physiology

### 2021-07-22 MS/ms.Rmd
Fixed Losia's comments

$^{1}$ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600

$^{2}$ Ecology & Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW

@@

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Such differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., $Q_{10}$) that provide comparative physiologists with a means to remove 'nuisance heterogeneity' without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive 'new' effects sizes and sampling variances opening up new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Meta-analysis has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology  [@Gurevitch2018; @Nakagawa2017; @Glass2015]. A meta-analysis provides quantitatively informed answers to research questions by synthesising research results across studies and identifying sources of variation across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Ideally, meta-analyses are part of a systematic review and carefully report upon this entire process to ensure the review and data collation are transparent and reproducible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021].  Aside from providing an up-to-date overview of a research field there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) we can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Meta-analyses have already provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis-phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Despite its widespread adoption and well-established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. For example, experimental designs in the field of comparative physiology can vary greatly in the temperatures or dosages of chemicals that they use in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal across studies. However, to many, having highly heterogeneous effects "...is the spice of life" [p.g., 519; @Cooperetal2019] because it provides opportunities to explore the reasons for why effects vary within and across studies [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Then, we present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We also formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate 'nuisance heterogeneity' at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (not just sampling variance) [@Nakagawa2017; @Borenstein2019]. There are many phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating the 'true' effect [@Borenstein2019]. In a meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known [@Gurevitch2018; @Koricheva2013]. While unweighted analyses are common, weighting studies makes overall estimates more precise, reliable and less susceptible to publication bias, even if the overall mean itself is unbiased [@Morrissey2016a]. As such, weighting effect size estimates by their inverse sampling variance can be crucial in meta-analysis. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! Some variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this to determine how experiments can be designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. While these factors are likely to have an impact on the magnitude of an effect, they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

```{r, fig1, fig.cap="*Sources of effect size heterogeneity in meta-analyses*. Heterogeneity can come in various forms, from effect measures varying as a result of shared evolutionary history (i.e., phylogeny), different trait types measured, and sources of 'nuisance heterogeneity'. In this example, nuisance heterogeneity is generated by differences in the experimental temperatures at which the animals were tested. However, nuisance heterogeneity can come in various other forms such as differences in dosage, or biochemical variables between used in the experiments. We can control for these types of heterogeniety by reformalising effect sizes and/or using multilevel metaregression as we describe in our review."}
image <- image_read(here::here("figs", "Fig1.png"))
image
```

Besides sampling variance, all the various sources of heterogeneity contribute (to varying extents) to the 'apples and oranges' problem often touted as compromising the reliability of meta-analyses. Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can indeed weaken the reliability of a meta-analysis. However, many argue that heterogeneity is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the founder of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. Whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples, then mixing fruit will not be ideal. However, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. By synthesizing diverse study outcomes, not only does it allow to report on overall effects (if the question is relevant to ask), but it allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches that can explain the diversity of effect outcomes observed. Doing so provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. Therefore, understanding overall effects in each of these groups separately, and assess the extent to which they differ, may be more useful.

There are, however, certain types of heterogeneity that transcend different questions or that are simply a 'nuisance' to interpreting effects. In other words, if we compare different varieties of 'apples', is it still fair to synthesize these studies to understand overall effects? In comparative physiology, this situation might be analogous to a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Here, we focus on meta-analytic tools to circumvent these issues and demonstrate that the ‘apples and oranges’ problem is not really a problem at all.

@@

Before embarking overviewing strategies to deal with 'nuisance heterogeneity', it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, but this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. We conducted a literature survey to better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled, we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

All searches were conducted in Scopus between May 16 - 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title and abstract, restricting our search to key comparative physiology journals (detailed in Table S1 and S2 - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided a random sample to inform on the questions we sought to address (see below).

We included comparative / meta-analyses that were focused on comparative physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. Quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers originally identified, 79 full texts were screened for eligibility, and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Information*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as 'raw'). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision about whether the effect measure was likely to be impacted. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured. The magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the pH or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected 'nuisance heterogeneity' to be common to most effect sizes. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multilevel meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species and observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models (including mixed effect models); and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models) [see @NakagawaSantos2012 for discussion of models]. Notably, only model types 1-5 are weighted models because they weight effect sizes based on the inverse of their sampling variance or sample size.

@@

# Maybe add some examples for each effect size in a new column. SHould help
eg <- c("CTmin, CTmax, Metabolic Rate (MR)",
        "Variability in CTmin, CTmax, MR",
        "Ratio between BPA treatment (BPA-exposed group) and control (no BPA)",
        "Difference in immune response between males and females or performance difference in the presence of stressor compared to absence of stressor",
        "Relationship between sex hormones and immune responses or metabolic rate and behaviour")

>>>>

$^{1}$ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia

$^{2}$ Ecology & Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW 2052, Australia

@@

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening up new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

Meta-analysis has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is increasingly prominent in the field of comparative physiology  [@Gurevitch2018; @Nakagawa2017; @Glass2015]. Meta-analyses answer research questions by synthesising research results and identifying sources of variation across studies [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Arnqvist1995; @Koricheva2013; @Borenstein2019]. Ideally, meta-analyses are part of a systematic review and methods are carefully reported upon to ensure the review and data collation are transparent and reproducible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolutionary Biology (PRISMA-EcoEvo) checklist, @ODea2021].

A meta-analysis can have three goals: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify research gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. Meta-analysing independent studies provides greater statistical power and precision than what any individual study on its own would be able to provide - particularly given that most empirical studies are already under-powered in many areas of biology [@Forstmeier2017; @Jennions2003; @Button2013]. By expressing study effects on a common scale (i.e., standardised effect size) we can gain broader insight into the direction and efficacy of a particular treatment effect or the strength of a relationship between two variables of interest [@Gurevitch2018; @Koricheva2013]. Meta-analyses have already provided comparative physiologists with powerful insights on pressing global challenges - from testing whether physiological plasticity can buffer organisms against climate change [e.g, @Seebacher2015] to the degree to which endocrine disrupting chemicals, such as bis-phenol A (BPA), impact aquatic organisms [e.g., @WuSeebacher2020].

Despite its widespread adoption and well-established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that are not comparable [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units [@Arnqvist1995; @Stewart2010]. For example, experimental designs in the field of comparative physiology can vary greatly in the temperatures or dosages of chemicals that they use in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal across studies. However, to many, having highly heterogeneous effects "...is the spice of life" [p.g., 519; @Cooperetal2019] because it provides opportunities to explore the reasons for why effects vary within and across studies [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015].

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. Then, we present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We also formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate 'nuisance heterogeneity' at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (not just sampling variance) [@Nakagawa2017; @Borenstein2019].

There are many phenomena driving effect heterogeneity. First, there is inherent uncertainty in estimating the ‘true’ population effect (e.g., a correlation coefficeiont) from smaller samples [@Borenstein2019]. A formal meta-analysis weights sample estimates by their precision, and removes this source of heterogeneity [@Gurevitch2018; @Koricheva2013]. Removing sampling variance is possible because we know, mathematically, how to calculate it for many effect measures. While unweighted analyses are common, weighting studies makes overall estimates more precise, and less susceptible to publication bias (see below), even if the overall mean itself is unbiased [@Morrissey2016a]. Second, there are sociological factors impacting heterogeneity. These include the ease with which novel and significant results are published relative to non-significant results, manifesting as publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, real biological processes can drive effect heterogeneity. This is particularly true in comparative physiology where we synthesise data from different species with varying life-histories, habitats, mating systems and reproductive modes. As comparative physiologists, this is the stuff that gets us up every morning! Some variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions.

Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this to determine how experiments can be designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. While these factors are likely to have an impact on the magnitude of an effect, they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

```{r, fig1, fig.cap="**Sources of effect size heterogeneity in meta-analyses**. Heterogeneity can come in various forms, from effect measures varying as a result of shared evolutionary history (i.e., phylogeny), the different trait types measured, and sources of 'nuisance heterogeneity'. In this example, nuisance heterogeneity is generated by differences in the experimental temperatures at which the animals were tested. However, nuisance heterogeneity can come in various other forms such as differences in dosage. We can control for these types of heterogeniety by reformalising effect sizes and/or using multilevel metaregression."}
image <- image_read(here::here("figs", "Fig1.png"))
image
```

Besides sampling variance, all the sources of heterogeneity discussed above contribute (to varying extents) to the 'apples and oranges' problem often touted as compromising the reliability of meta-analyses. Combining standardised effects, while ignoring these sources of heterogeneity can indeed weaken the reliability of a meta-analysis. However, many argue that heterogeneity is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the founder of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. Whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in generalising only to apples, then mixing fruit will not be ideal. However, if we want to understand why different 'fruit' vary in their effects, then we need to consider apples and oranges together [@Borenstein2019].

We tend to agree with @Glass2015, and his sentiment is particularly relevant to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. Synthesizing diverse study outcomes allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches (e.g., short-term vs long-term studies) that can explain the diversity of effect outcomes observed. Doing so provides a rich set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. It may, therefore be more useful tounderstand overall effects in each of these groups separately, and assess the extent to which they differ.

Certain types of heterogeneity, however, get in the way of understanding real biological patterns and are simply a 'nuisance' to interpreting effects. In other words, if we compare different varieties of 'apples', is it still fair to synthesize these studies to understand overall effects? In comparative physiology, this situation might be analogous to a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Here, we focus on meta-analytic tools to circumvent these issues and demonstrate that the ‘apples and oranges’ problem is not really a problem at all.

@@

Before overviewing strategies to deal with nuisance heterogeneity, it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As such, we conducted a literature survey to better understand the different types of effect sizes being used in comparative physiology, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled. After searching for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species, we we asked: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was such variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

All searches were conducted in Scopus between May 16 and 18, 2021. Briefly, we first did a series of pilot searches (detailed in Supplementary Materials) to refine our search query [See @Foo2021 for an overview of this process]. We searched for studies that included: 'meta-analysis', 'meta regression' , 'comparative analysis', 'comprehensive analysis' 'global analysis' or 'macro physiology' in their title or abstract, restricting our search to key comparative physiology journals (detailed in Table S1 and S2 - Supplementary Materials) within the last 6 years. We expected this to provide a contemporary overview of quantitative syntheses in comparative physiology while making the survey and screening feasible. We acknowledge that this may have missed some important, and influential, quantitative syntheses in the field of comparative physiology, but likely provided an informative random sample.

We included comparative studies / meta-analyses that were focused on physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data, such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. We excluded quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics. There were also meta-analyses that focused on colour evolution which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

All title, abstract and full text screening was done by two people (DWAN and ML). Any disagreements were resolved by discussion to reach a consensus. Out of the 417 papers originally identified, 79 full texts were screened for eligibility, and we included a total of 62 for final data extraction. For full details on the search strings used, search dates, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the Supplementary Information.

To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (*lnRR*), Standardised Mean Differences (*SMD*; e.g., Hedges *d*, *g* or Cohen's *d*), Correlation or Fisher's Z -transformed correlation (*Zr*), Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we labelled as 'raw'). All other effects were labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision about whether the effect measure was likely to be impacted. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured. The magnitude of differences between two treatments (e.g, contrast-based effect sizes -- *SMD* or *lnRR*) might depend on the pH or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected 'nuisance heterogeneity' to be common to most effect sizes. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random-effects meta-analysis (i.e., intercept-only model, but has random study effects); 3) multilevel meta-analysis (i.e., intercept-only model, but has multiple random effect levels such as study, species and observation-level random effects); 4) meta-regression models (i.e., a multilevel or random-effects model with fixed effects - including random effects); 5) weighted regression models (including mixed- effects models); and 6) linear regression models (including linear and generalised linear mixed-effect models, and phylogenetic least squares models) [see @NakagawaSantos2012 for discussion of models]. Notably, only model types 1--5 are weighted models because they weight effect sizes based on the inverse of their sampling variance or sample size.

@@

# Maybe add some examples for each effect size in a new column. SHould help
eg <- c("CTmin, CTmax, Metabolic Rate (MR)",
        "Variability in CTmin, CTmax, Metabolic Rate (MR)",
        "Ratio between BPA treatment (BPA-exposed group) and control (no BPA)",
        "Difference in immune response between males and females or performance difference in the presence of stressor compared to absence of stressor",
        "Relationship between sex hormones and immune responses or metabolic rate and behaviour")
